Machine Learning for Mechanistic Models of Metapopulation Dynamics

### 2023-11-12 ms.Rnw
arxiv version

\documentclass[12pt]{article}

\newcommand\arxiv[2]{#1}  %% for arxiv version
%% \newcommand\arxiv[2]{#2}  %% suppressing arxiv formatting

@@

\topmargin 0.0cm
\oddsidemargin 0.2cm
\textwidth 16cm 
\textheight 21cm
\footskip 1.0cm
\usepackage{xcolor}
%orange for EI
\definecolor{orange}{rgb}{1,0.5,0}
\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
\newcommand\eic[1]{\textcolor{orange}{[#1]}}
%green for PN
\definecolor{green}{rgb}{0,0.5,0}
\newcommand\pn[2]{\sout{#1} \textcolor{green}{#2}}
\newcommand\pnc[1]{\textcolor{green}{[#1]}}
%purple for JF
\definecolor{purple}{rgb}{0.5,0,1}
\newcommand\jf[2]{\sout{#1} \textcolor{purple}{#2}}
\newcommand\jfc[1]{\textcolor{purple}{[#1]}}
%teal for TODO
\definecolor{teal}{rgb}{0,0.5,0.8}
\newcommand\todo[1]{\textcolor{teal}{[TO DO? #1]}}

@@

\begin{abstract}
Mathematical models in ecology and epidemiology must be consistent with observed data in order to generate reliable knowledge and sound policy. Metapopulation systems, which consist of a collection of sub-populations at various locations, pose technical challenges in statistical inference due to nonlinear, stochastic interactions. Difficulties encountered in these methodological issues can obstruct the core scientific questions concerning the link between the mathematical models and the data. Progress in statistically efficient simulation-based inference for partially observed stochastic dynamic systems has enabled the development of statistically rigorous approaches to the analysis of nonlinear but low-dimensional systems. Recently, an algorithm has been developed which enables comparable inference for higher-dimensional models arising in metapopulation systems. The COVID-19 pandemic provides a situation where mathematical models and their policy implications were widely visible, and we revisit an influential metapopulation model used to inform basic epidemiological understanding early in the pandemic. Our methods support self-critical data analysis, enabling us to identify and address weaknesses, and leading to a new model with substantially improved statistical fit and parameter identifiability. Our results suggest that the lockdown initiated on January 23, 2020 in China was more effective than previously thought. We proceed to recommend statistical analysis standards for future metapopulation system modeling.
\end{abstract}

@@

Matching summary statistics of simulations to the corresponding data statistic is, in principle, a readily applicable inference approach for a wide class of models including metapopulation models.
In the context of Bayesian inference this is called Approximate Bayesian Computing \cite{conlan12}.
However, informative, low-dimensional summary statistics can be hard to construct even for low-dimensional nonlinear systems.
This can make summary statistic methods statistically inefficient \cite{fasiolo16,shrestha11}.

@@

For low-dimensional systems, particle filter methods are applicable to a flexible class of models for low-dimensional nonlinear systems \cite{breto09,king16}.
They have the attractive feature of statistically efficient use of data, i.e., accessing the likelihood function for Bayesian or likelihood-based inference, with approximation resulting only from finite Monte Carlo effort.
For high-dimensional systems, scalability considerations demand further approximations since particle filters suffer acutely from the ``curse of dimensionality'' \cite{bengtsson08}.
In contrast, block particle filter (BPF) methods achieve scalability by updating particles through localized resampling, rather than employing the linear update of EnKF \cite{rebeschini15}.
It is an empirical question which of these approximations is more successful on metapopulation models,  with prior evidence favoring BPF \cite{ionides21}.
In this paper, we not only aim to compare these methods in the context of metapopulation modeling but importantly also we reveal issues of model assessment when fitting complicated models to large datasets.
We introduce a model diagnosis procedure to mitigate the risk of selecting an inappropriate model, an essential step in informing public policy decisions.

@@

Our starting point is model {\LiMobility} which is based on the model of \cite{li20} and is fully described in Supplementary Sec.~\suppSecLiModel.
We consider the full dataset, from January 10 to February 8, with transmission parameters re-estimated following the lockdown on January 23; these correspond to the periods~1 (January 10 to January 22) and 3 (January 24 to February 8) of \cite{li20}.
Some minor differences between {\LiMobility} and \citep{li20} were introduced to enable us to place their model within the general metapopulation framework of spatiotemporal partially observed  continuous-time Markov processs described by \citep{ionides21}.
Despite these modifications, simulations from {\LiMobility}, using the parameters of \citep{li20}, closely match simulations from the code provided by \cite{li20} (Supplementary Sec.~\suppSecLiModel).
However, inspection of the mobility data reveals that some small cities have no recorded incoming travelers, and therefore no possibility of a SARS-CoV2 introduction within {\LiMobility} (or the model of \cite{li20}) (Supplementary Sec.~\suppSecMobility).
This minor limitation formally results in a likelihood of zero for {\LiMobility} (i.e., it is impossible for the simulation model to reproduce the observed spatiotemporal dataset), and hence a log-likelihood of $-\infty$.

@@

All parameter estimates are reported in Supplementary Table~\suppTableResults.
Here, we discuss the estimated basic reproductive number (i.e., the expected number of secondary infections from one index case in a fully susceptible population), denoted by $\Rzero^{\before}$ and $\Rzero^{\after}$ before and after the January 23rd lockdown.
Our estimates for model {\RevisedModelConstrained}, are $\Rzero^{\before}=\Sexpr{metapoppkg::myround(ci_c["R0_be", "mle"],2)}$ with confidence interval (CI) (\Sexpr{metapoppkg::myround(ci_c["R0_be", "ci1"],2)},\Sexpr{metapoppkg::myround(ci_c["R0_be", "ci2"],2)}) and $\Rzero^{\after}=\Sexpr{metapoppkg::myround(ci_c["R0_af","mle"],2)}$  with CI (\Sexpr{metapoppkg::myround(ci_c["R0_af", "ci1"],2)},\Sexpr{metapoppkg::myround(ci_c["R0_af","ci2"],2)}), where the estimates and their associated 95\% CIs obtained by profile likelihood (Supplementary Sec.~\suppSecInference).
This implies that the Chinese government non-pharmaceutical interventions instituted on and around January 23 reduced $\Rzero$ by a factor of $\Sexpr{metapoppkg::myround(ci_c["R0_be","mle"]/ci_c["R0_af","mle"],1)}$.
By contrast, the estimates of \cite{li20} are $\widetilde\Rzero^{\before}=2.38$  with CI $(2.03,2.77)$ and $\widetilde\Rzero^{\after}=0.98$  with CI $(0.83,1.16)$, implying reduction by a factor of $\Sexpr{metapoppkg::myround(2.38/0.98,1)}$.
For comparison, interventions implemented across a panel of 41 countries (34 European) were estimated to reduce $\Rzero$ by a factor of $\Sexpr{metapoppkg::myround(1/(1-0.77),1)}$  with CI $(\Sexpr{metapoppkg::myround(1/(1-0.65),1)},\Sexpr{metapoppkg::myround(1/(1-0.85),1)})$ \citep{brauner21}.
Our estimate for $\Rzero$ before lockdown is toward the high end of comparable estimates, reviewed by \citep{park20-jcm}.
An alternative metaopopulation analysis of the pre-lockdown China data, with a deterministic transmission model, obtained an $\Rzero$ estimate of 3.11  with CI  (2.39,4.13) \citep{read21}.

@@

Likelihood-based inference via particle filters has been considered inaccessible for metapopulation models due to the ``curse of dimensionality'' \cite{bengtsson08}.
However, block particle filter methods can be effective on metapopulation models, as demonstrated in this paper and previously \cite{ionides21,ning23,wheeler23}.
All high-dimensional nonlinear filters entail numerical approximation, and these can be assessed by comparing predictive skill (i.e., the estimated log-likelihood) between different filters.
The ensemble Kalman filter provides a suitable point of comparison, since it has excellent scalability properties, modest capability to handle nonlinearities, and has been demonstrated on various epidemiological systems \cite{shaman12,yang14,yang15,pei18,yang21,kramer20-plos-cb,cascante-vega22}.

Software environments are critical for providing data analysis that is not only reproducible but also readily extendable.
Scientists developing a data analysis should develop an environment that empowers them to explore their own models and data, and then they should share this environment as part of the publication process.
In practice, this involves encapsulating data analysis within a software package  that immerses the user in a documented environment where the models, methods and data used for the article can be readily be experimented with.
Trustworthy data analysis should be supported by unit testing and documentation, and the quality of this support should be one of the considerations in evaluation of the data analysis.
In other words, the article presenting the research should be part of a compendium \cite{gentleman07}.
The compendium for this article is comprised of the article source code, at \url{https://github.com:jifanli/metapop_article}, together with the software environment for the data analysis, provided by the R package at \url{https://github.com:jifanli/metapoppkg}.

Our research demonstrates that techniques proven effective in low-dimensional systems, such as population dynamics at one or two locations, can be extended to address larger metapopulation systems. This extension allows us to leverage well-established best practices from time series analysis, leading to a statistically principled approach. This approach enables us to identify and rectify model weaknesses that might otherwise remain undetected.
Failure to address these weaknesses can lead to issues of irreproducibility and the provision of suboptimal policy recommendations when developing models for complex dynamic systems \cite{saltelli20,ioannidis22}.
Principles of good data analysis for population dynamics are presumably similar to general principles of data science \cite{yu20} but require some adaptation to the specific situation.
Here, we build on \cite{saltelli20,ioannidis22,yu20}, by demonstrating the feasibility and desirability of metapopulation analysis meeting the specific set of criteria outlined below:

@@

\noindent {\it Model}.
Our model is a spatiotemporal partially observed Markov process (SpatPOMP) \citep{ionides21}.
A partially observed Markov process (POMP) \citep{breto09,king16} is a stochastic dynamic model comprised of: (i) a latent process having the Markov property that the future is conditionally independent of the past given the present; (ii) a measurement process, describing how the data are modeled as noisy and incomplete observations of the latent process.
The SpatPOMP framework adds an extra assumption that the latent process is comprised of a collection of units, each of which has its own latent process and observation process.
The latent processes for each unit can be interdependent, but the observations for a given unit are required to depend only on the latent process for that unit.
In our context, each city is a unit.
General notation for SpatPOMP models is set up in Supplementary Sec.~\suppSecOurModel, and this notation is subsequently used to define mathematically the specific SpatPOMP models studied in this paper ({\LiMobility}, {\LiParams},{\RevisedModelUnconstrained} and {\RevisedModelConstrained}).
All the models under consideration have an SEAIR structure, as described in Figure~\ref{fig:flow_diagram}.

@@

\bibliography{bib-metapop}

>>>>

%\documentclass[12pt]{article}
\documentclass[11pt]{article} \usepackage{fullpage}
\pdfoutput=1 
\newcommand\arxiv[2]{#1}  %% for arxiv version
%% \newcommand\arxiv[2]{#2}  %% suppressing arxiv formatting

@@

\usepackage{xcolor}
%orange for EI
\definecolor{orange}{rgb}{1,0.5,0}
\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
\newcommand\eic[1]{\textcolor{orange}{[#1]}}
%green for PN
\definecolor{green}{rgb}{0,0.5,0}
\newcommand\pn[2]{\sout{#1} \textcolor{green}{#2}}
\newcommand\pnc[1]{\textcolor{green}{[#1]}}
%purple for JF
\definecolor{purple}{rgb}{0.5,0,1}
\newcommand\jf[2]{\sout{#1} \textcolor{purple}{#2}}
\newcommand\jfc[1]{\textcolor{purple}{[#1]}}
%teal for TODO
\definecolor{teal}{rgb}{0,0.5,0.8}
\newcommand\todo[1]{\textcolor{teal}{[TO DO? #1]}}

@@

\begin{abstract}
Mathematical models in ecology and epidemiology must be consistent with observed data in order to generate reliable knowledge and sound policy. Metapopulation systems, which consist of a collection of sub-populations at various locations, pose technical challenges in statistical inference due to nonlinear, stochastic interactions. Difficulties encountered in these methodological issues can obstruct the core scientific questions concerning the link between the mathematical models and the data. Progress in statistically efficient simulation-based inference for partially observed stochastic dynamic systems has enabled the development of statistically rigorous approaches to the analysis of nonlinear but low-dimensional systems. Recently, an algorithm has been developed which enables comparable inference for higher-dimensional models arising in metapopulation systems. The COVID-19 pandemic provides a situation where mathematical models and their policy implications were widely visible, and we revisit an influential metapopulation model used to inform basic epidemiological understanding early in the pandemic. Our methods support self-critical data analysis, enabling us to identify and address model limitations, and leading to a new model with substantially improved statistical fit and parameter identifiability. Our results suggest that the lockdown initiated on January 23, 2020 in China was more effective than previously thought. We proceed to recommend statistical analysis standards for future metapopulation system modeling.
\end{abstract}

@@

Matching summary statistics of simulations to the corresponding data statistic is, in principle, a readily applicable inference approach for a wide class of models including metapopulation models.
In the context of Bayesian inference this is called Approximate Bayesian Computing \cite{conlan12}.
However, informative, low-dimensional summary statistics can be hard to construct even for low-dimensional nonlinear systems.
This can make summary statistic methods statistically inefficient \cite{fasiolo16}.

@@

For low-dimensional systems, particle filter methods are applicable to a flexible class of models for low-dimensional nonlinear systems \cite{breto09,king16}.
They have the attractive feature of statistically efficient use of data, i.e., accessing the likelihood function for Bayesian or likelihood-based inference, with approximation resulting only from finite Monte Carlo effort.
For high-dimensional systems, scalability considerations demand further approximations since particle filters suffer acutely from the ``curse of dimensionality'' \cite{bengtsson08}.
In contrast, block particle filter (BPF) methods achieve scalability by updating particles through localized resampling, rather than employing the linear update of EnKF \cite{rebeschini15}.
It is an empirical question which of these approximations is more successful on metapopulation models,  with prior evidence favoring BPF \cite{ionides23-jasa}.
In this paper, we not only aim to compare these methods in the context of metapopulation modeling but importantly also we reveal issues of model assessment when fitting complicated models to large datasets.
We introduce a model diagnosis procedure to mitigate the risk of selecting an inappropriate model, an essential step in informing public policy decisions.

@@

Our starting point is model {\LiMobility} which is based on the model of \cite{li20} and is fully described in Supplementary Sec.~\suppSecLiModel.
We consider the full dataset, from January 10 to February 8, with transmission parameters re-estimated following the lockdown on January 23; these correspond to the periods~1 (January 10 to January 22) and 3 (January 24 to February 8) of \cite{li20}.
Some minor differences between {\LiMobility} and \citep{li20} were introduced to enable us to place their model within the general metapopulation framework of spatiotemporal partially observed  continuous-time Markov processs described by \citep{ionides23-jasa}.
Despite these modifications, simulations from {\LiMobility}, using the parameters of \citep{li20}, closely match simulations from the code provided by \cite{li20} (Supplementary Sec.~\suppSecLiModel).
However, inspection of the mobility data reveals that some small cities have no recorded incoming travelers, and therefore no possibility of a SARS-CoV2 introduction within {\LiMobility} (or the model of \cite{li20}) (Supplementary Sec.~\suppSecMobility).
This minor limitation formally results in a likelihood of zero for {\LiMobility} (i.e., it is impossible for the simulation model to reproduce the observed spatiotemporal dataset), and hence a log-likelihood of $-\infty$.

@@

All parameter estimates are reported in Supplementary Table~\suppTableResults.
Here, we discuss the estimated basic reproductive number (i.e., the expected number of secondary infections from one index case in a fully susceptible population), denoted by $\Rzero^{\before}$ and $\Rzero^{\after}$ before and after the January 23rd lockdown.
The mathematical formula used to calculate  $\Rzero^{\before}$ and $\Rzero^{\after}$, in terms of the model parameters, is provided in Supplementary Table~\suppTableResults.
Our estimates for model {\RevisedModelConstrained}, are $\Rzero^{\before}=\Sexpr{metapoppkg::myround(ci_c["R0_be", "mle"],2)}$ with confidence interval (CI) (\Sexpr{metapoppkg::myround(ci_c["R0_be", "ci1"],2)},\Sexpr{metapoppkg::myround(ci_c["R0_be", "ci2"],2)}) and $\Rzero^{\after}=\Sexpr{metapoppkg::myround(ci_c["R0_af","mle"],2)}$  with CI (\Sexpr{metapoppkg::myround(ci_c["R0_af", "ci1"],2)},\Sexpr{metapoppkg::myround(ci_c["R0_af","ci2"],2)}), where the estimates and their associated 95\% CIs obtained by profile likelihood (Supplementary Sec.~\suppSecInference).
This implies that the Chinese government non-pharmaceutical interventions instituted on and around January 23 reduced $\Rzero$ by a factor of $\Sexpr{metapoppkg::myround(ci_c["R0_be","mle"]/ci_c["R0_af","mle"],1)}$.
By contrast, the estimates of \cite{li20} are $\widetilde\Rzero^{\before}=2.38$  with CI $(2.03,2.77)$ and $\widetilde\Rzero^{\after}=0.98$  with CI $(0.83,1.16)$, implying reduction by a factor of $\Sexpr{metapoppkg::myround(2.38/0.98,1)}$.
For comparison, interventions implemented across a panel of 41 countries (34 European) were estimated to reduce $\Rzero$ by a factor of $\Sexpr{metapoppkg::myround(1/(1-0.77),1)}$  with CI $(\Sexpr{metapoppkg::myround(1/(1-0.65),1)},\Sexpr{metapoppkg::myround(1/(1-0.85),1)})$ \citep{brauner21}.
Our estimate for $\Rzero$ before lockdown is toward the high end of previous estimates based on data up to February 2020, reviewed by \citep{park20-jcm}.
A contributing factor to our relatively high $\Rzero$ estimate may be that our mechanistic model includes transmission from asymptomatic individuals in the $\Rzero$ estimate \cite{subramanian21}.
An alternative metaopopulation analysis of the pre-lockdown China data, with a deterministic transmission model, obtained an $\Rzero$ estimate of 3.11  with CI  (2.39,4.13) \citep{read21}.

@@

Likelihood-based inference via particle filters has been considered inaccessible for metapopulation models due to the ``curse of dimensionality'' \cite{bengtsson08}.
However, block particle filter methods can be effective on metapopulation models, as demonstrated in this paper and previously \cite{ionides23-jasa,ning23,wheeler23}.
All high-dimensional nonlinear filters entail numerical approximation, and these can be assessed by comparing predictive skill (i.e., the estimated log-likelihood) between different filters.
The ensemble Kalman filter provides a suitable point of comparison, since it has excellent scalability properties, modest capability to handle nonlinearities, and has been demonstrated on various epidemiological systems \cite{shaman12,yang14,yang15,pei18,yang21,kramer20-plos-cb,cascante-vega22}.

Software environments are critical for providing data analysis that is not only reproducible but also readily extendable.
Scientists developing a data analysis should develop an environment that empowers them to explore their own models and data, and then they should share this environment as part of the publication process.
In practice, this involves encapsulating data analysis within a software package  that immerses the user in a documented environment where the models, methods and data used for the article can be readily be experimented with.
Trustworthy data analysis should be supported by unit testing and documentation, and the quality of this support should be one of the considerations in evaluation of the data analysis.
In other words, the article presenting the research should be part of a compendium \cite{gentleman07}.
The compendium for this article is comprised of the article source code, at \url{https://github.com/jifanli/metapop_article}, together with the software environment for the data analysis, provided by the R package at \url{https://github.com/jifanli/metapoppkg}.

Our research demonstrates that techniques proven effective in low-dimensional systems, such as population dynamics at one or two locations, can be extended to address larger metapopulation systems. This extension allows us to leverage well-established best practices from time series analysis, leading to a statistically principled approach. This approach enables us to identify and rectify model limitations that might otherwise remain undetected.
Failure to address these weaknesses can lead to issues of irreproducibility and the provision of suboptimal policy recommendations when developing models for complex dynamic systems \cite{saltelli20,ioannidis22}.
Principles of good data analysis for population dynamics are presumably similar to general principles of data science \cite{yu20} but require some adaptation to the specific situation.
Here, we build on \cite{saltelli20,ioannidis22,yu20}, by demonstrating the feasibility and desirability of metapopulation analysis meeting the specific set of criteria outlined below:

@@

\noindent {\it Model}.
Our model is a spatiotemporal partially observed Markov process (SpatPOMP) \citep{ionides23-jasa}.
A partially observed Markov process (POMP) \citep{breto09,king16} is a stochastic dynamic model comprised of: (i) a latent process having the Markov property that the future is conditionally independent of the past given the present; (ii) a measurement process, describing how the data are modeled as noisy and incomplete observations of the latent process.
The SpatPOMP framework adds an extra assumption that the latent process is comprised of a collection of units, each of which has its own latent process and observation process.
The latent processes for each unit can be interdependent, but the observations for a given unit are required to depend only on the latent process for that unit.
In our context, each city is a unit.
General notation for SpatPOMP models is set up in Supplementary Sec.~\suppSecOurModel, and this notation is subsequently used to define mathematically the specific SpatPOMP models studied in this paper ({\LiMobility}, {\LiParams},{\RevisedModelUnconstrained} and {\RevisedModelConstrained}).
All the models under consideration have an SEAIR structure, as described in Figure~\ref{fig:flow_diagram}.

@@

\bibliography{bib-metapop}
%\input{ms.bbl}
